---
title: "pr_housing_vacancy"
author: "Jorge R. Soldevila Irizarry"
date: "2026-03-03"
output: github_document
---

---
title: "Puerto Rico Housing Vacancy Analysis (2024)"
author: "Jorge R. Soldevila Irizarry"
date: "2026-03-03"
output: github_document
---

# Overview
This script retrieves housing vacancy data for Puerto Rico from the US Census Bureau (ACS 5-Year Estimates, 2024). It pulls data at two geographic levels:
1. **Municipio (County)**
2. **Census Tract**

## Setup and Libraries

```{r Call library packages, warning=FALSE, message=FALSE}
# Install pacman if not already installed
if (!require("pacman")) install.packages("pacman")

# Load libraries
pacman::p_load(
  tidyverse,       # Data manipulation (dplyr, readr, etc.)
  tidycensus,      # Interface with US Census API
  sf               # Handling geospatial data (geometry = TRUE)
)

# Note: Ensure your Census API key is installed.
# If not, run: census_api_key("YOUR_KEY_HERE", install = TRUE)
```

##Variable set up

```{r Variable Dictionary, warning=FALSE, message=FALSE}

# Variable dictionary
housing_vars <- c(
  total_pop       = "B01003_001E",
  # Vacancy Status Breakdown (B25004)
  total_vac_det   = "B25004_001E",
  vac_rent        = "B25004_002E",
  vac_rent_no_occ = "B25004_003E",
  vac_sale        = "B25004_004E",
  vac_sold        = "B25004_005E",
  vac_recr        = "B25004_006E", # For seasonal/recreational use
  vac_migr        = "B25004_007E", # For migrant workers
  vac_oth         = "B25004_008E",
  # Tenure (B25003)
  total_occ_ten   = "B25003_001E",
  occ_own         = "B25003_002E",
  occ_rent        = "B25003_003E",
  # General Occupancy (B25002)
  tot_hous        = "B25002_001E",
  tot_occ         = "B25002_002E",
  tot_vac         = "B25002_003E"
)

```

##County level data

```{r Housing by county, message=FALSE, warning=FALSE}

pr_cnty_df <- get_acs(
  geography = "county",
  variables = housing_vars,
  state = "PR",
  year = 2024,
  survey = "acs5",
  output = "wide",
  geometry = TRUE,
  keep_geo_vars = TRUE,
  progress_bar = FALSE) %>%
  select(-ends_with("M")) %>% # Removing Margin of Error (MOE) columns for cleaner view
  mutate(
    land_area_sqmi = ALAND / 2589988.10,
    pop_density = total_pop / land_area_sqmi,
    pct_occ = tot_occ / tot_hous * 100,
    pct_vac = tot_vac / tot_hous * 100,
    pct_rent_occ = occ_rent / total_occ_ten * 100,
    pct_seasonal = vac_recr / total_vac_det * 100,
    pct_oth_vac = vac_oth / total_vac_det * 100
  )

st_write(pr_cnty_df, "pr_county_housing.shp", delete_dsn = TRUE)
```

##Tract level data

```{r Housing by tract, message=FALSE, warning=FALSE}

pr_tract_df <- get_acs(
  geography = "tract",
  variables = housing_vars,
  state = "PR",
  year = 2024,
  survey = "acs5",
  output = "wide",
  geometry = TRUE,
  keep_geo_vars = TRUE,
  progress_bar = FALSE) %>%
  select(-ends_with("M")) %>% # Removing Margin of Error (MOE) columns for cleaner view
   mutate(
    land_area_sqmi = ALAND / 2589988.10,
    pop_density = total_pop / land_area_sqmi,
    pct_occ = tot_occ / tot_hous * 100,
    pct_vac = tot_vac / tot_hous * 100,
    pct_rent_occ = occ_rent / total_occ_ten * 100,
    pct_seasonal = vac_recr / total_vac_det * 100,
    pct_oth_vac = vac_oth / total_vac_det * 100
  )

st_write(pr_tract_df, "pr_tract_housing.shp", delete_dsn = TRUE)
```


